\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2025-05-25 misc/JSS article/final/article.Rnw
Minors.

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://sebkrantz.github.io/collapse/. Linecount (v2.1.2): \proglang{R}: 13,785, \proglang{C}: 19,013, \proglang{C++}: 9,844. \hphantom{aaa.}Exported namespace: 392 objects, of which 238 functions (excl. methods and shorthands), and 2 datasets.}
% Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
Its core functionality includes a rich set of S3 generic (grouped, weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}). It also provides powerful data manipulation functions, such as vectorized and verbose hash-joins or fast (aggregation, recast) pivots, functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools. This functionality is supported by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels. It also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data. \pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations for vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}. It is also globally and interactively configurable, allowing changing defaults of key function arguments, such as the \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifying the package namespace to mask equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents. See Section~\ref{sec:glob_opt}.\vspace{-5mm}} % \newline

>>>>

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://sebkrantz.github.io/collapse/. Linecount (v2.1.2): \proglang{R}: 13,785, \proglang{C}: 19,013, \proglang{C++}: 9,844. \hphantom{aaa.}Exported namespace: 392 objects, of which 238 functions (excl. methods and shorthands), and 2 datasets.}
% Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
Its core functionality includes a rich set of S3 generic (grouped, weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}). It also provides powerful data manipulation functions, such as vectorized and verbose hash-joins or fast (aggregation, recast) pivots, functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools. This functionality is supported by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels. It also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data. \pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations for vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}. It is globally and interactively configurable, allowing changes to defaults of key function arguments, such as the \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modification of the package namespace to mask equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents. See Section~\ref{sec:glob_opt}.\vspace{-5mm}} % \newline
